---
title: "Step 3: ATAC-Seq-Data-Analysis-QC on all cells"
subtitle: "Step 3.2.2: Visualisation of ATAC-Seq-Data-Analysis-QC on all cells using seurat's 'LogNormalize'"
date: "2025-02-11"
author: Xenia Sirodzha
output:
  html_document: 
    toc: yes
    toc_depth: 4
    toc_float: yes
    df_print: paged
---

As the workflow for ATAC-seq QC was established in part 1, I performed this workflow on the entire data set. However, performing pseudo-bulking and normalization and subsequent visualization with the data from all cells is very computationally intensive. Therefore, visualization was performed separately from the pseudo-bulking and normalization here. 

# Libraries

These are the packages required for the code below.
```{r libraries}
library(tidyverse) 
library(binom)
library(lubridate)
library(broom)
library(scales) #for scale transformations
library(cowplot) #arranging plots 
library(ggpubr)
library(ggridges)
library(ggthemes) #stylising plots
library(ggplot2)
library(ggsci) #color palettes
library(readxl) #for loading excel files
library(Seurat) #for rna-seq data, atac-seq data
library(Signac) #for atac seq data
```

# 1 Data Loading

For visualization, the manually normalized data set from "02-11-ATAC-Seq-Analysis-Step-3.1.2-seurat on all celltypes" was loaded into the session.
```{r data loading}
readRDS("Data/Lin_2023_zfish_snATAC-seq/Cell-type_Peak_Matrix.rds") -> zfish_snATAC_seq_pk_mtrx
readRDS("Data/psd.bulk.zfish_atac.rds") -> psd.bulk.zfish_atac
```


# 2 Data wrangling for plotting

After normalisation, I prepared the data to plot with tidyverse. First, I converted the matrix to a data frame, then I calculated the values which will be plotted as well(sum of pk_counts, coefficient of variance (cv))

```{r data wrangling for visualization}
as.data.frame(psd.bulk.zfish_atac@assays$atac$data)%>% 
  pivot_longer(cols=!contains("rownames"), ,values_to = "pk_counts", names_to = "celltype") %>% 
  mutate(pk_count_sum= sum(pk_counts)) %>% 
  mutate(cv=sd(pk_counts) / mean(pk_counts)) %>% 
  group_by(celltype) %>% 
  mutate(pk_count_sum_celltype= sum(pk_counts)) %>% 
  mutate(cv_celltype=sd(pk_counts) / mean(pk_counts)) %>% 
  mutate(all="all celltypes")-> df_psd.bulk.zfish_atac

df_psd.bulk.zfish_atac
```

# 3 Plotting

Subsequently, the wrangled data was visualized analogous to the script in step 2 "02-06-ATAC-Seq-Analysis-Step-2.1 seurat norm". 

## 3.1 Distribution after pseudo-bulking by celltype

Then, the distribution of peaks overall and by celltype was visualized below.
```{r histograms counts}
# overall
df_psd.bulk.zfish_atac %>% 
 ggplot(aes(x=pk_counts))+geom_histogram(binwidth = 0.02, color="gray15", alpha=0.8)+
 scale_y_continuous(breaks = scales::breaks_pretty(n=10))+
   scale_x_continuous(breaks = scales::breaks_pretty(n=4))+
  coord_fixed(ratio = 0.0000015)+
  theme_bw()+ theme(text = element_text(size = 9, family="sans", color="black"), #stylisation
        axis.text = element_text(size=9, color="black"),
        axis.title=element_text(size=9, face = "bold"),
        plot.subtitle=element_text(size=10.5),
        plot.title=element_text(size=11, face = "bold"), 
        legend.position ="none") ->dis.counts.1

# by celltype
 ggplot(df_psd.bulk.zfish_atac, aes(x=pk_counts, fill = celltype))+
  geom_histogram(binwidth = 0.02, color="gray15", alpha=0.8)+
  scale_y_continuous(breaks = scales::breaks_pretty(n=5))+
   scale_x_continuous(breaks = scales::breaks_pretty(n=4))+
  facet_wrap(facets = "celltype")+
  theme_bw()+ theme(text = element_text(size = 9, family="sans", color="black"), #stylisation
        axis.text = element_text(size=9, color="black"),
        axis.title=element_text(size=9, face = "bold"),
        plot.subtitle=element_text(size=10.5),
        plot.title=element_text(size=11, face = "bold"), 
        legend.position ="none") ->dis.counts.cell.1

 # display
dis.counts.1|dis.counts.cell.1
```

Overall, the counts are shifted towards zero and not normally distributed, indicating that seurat's LogNormalize did not normalize the data sufficiently. This also applies to the distributions per cell type.

## 3.3 Sum of all peaks

The sum of all peaks is 220,551.
```{r extracting value sum of all peaks}
unique(df_psd.bulk.zfish_atac$pk_count_sum)
```

This is 100 times less than with the manual method performed in "02-11-ATAC-Seq-Analysis-Step-3.1.1-log 1+x on all celltypes". 

Then, the sum of all peaks overall and by celltype was visualized below.
```{r bar plot sum of all peaks}

df_psd.bulk.zfish_atac %>% 
  ungroup() %>% 
  distinct(all, pk_count_sum) %>% 
  ggplot(aes(x=all, y=pk_count_sum))+
  geom_col(alpha=0.8, color="gray10")+
  coord_fixed(ratio = 0.00005)+
  theme_bw()+ theme(text = element_text(size = 9, family="sans", color="black"), #stylisation
        axis.text = element_text(size=9, color="black"),
        axis.title=element_text(size=9, face = "bold"),
        plot.subtitle=element_text(size=10.5),
        plot.title=element_text(size=11, face = "bold"), 
        legend.position ="none") -> sum.peaks.all.1


df_psd.bulk.zfish_atac %>% 
  group_by(celltype) %>% 
  distinct(pk_count_sum_celltype, pk_count_sum) %>% 
  ggplot(aes(x=celltype, y=pk_count_sum_celltype, fill=celltype))+
  geom_col(alpha=0.8, color="gray10")+
  theme_bw()+ theme(text = element_text(size = 9, family="sans", color="black"), #stylisation
        axis.text = element_text(size=9, color="black"),
        axis.title=element_text(size=9, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.subtitle=element_text(size=10.5),
        plot.title=element_text(size=11, face = "bold"), 
        legend.position ="none") -> sum.peaks.celltype.1

sum.peaks.all.1|sum.peaks.celltype.1
```

The sum of all peaks per cell type is almost identical (9000). This indicated that the transformation performed by seurat's "LogNormalize" does not preserve the underlying distribution of the values. 

## 3.4 Coefficient of variation 

Next, the coefficient of variation was displayed in a barplot. 
```{r bar plot coefficient of variation}
df_psd.bulk.zfish_atac %>% 
  ungroup() %>% 
  distinct(all, cv) %>% 
  ggplot(aes(x=all, y=cv ))+
  geom_col(alpha=0.8, color="gray10")+
  scale_y_continuous(n.breaks=10)+
  coord_fixed(ratio = 7)+
  theme_bw()+ theme(text = element_text(size = 9, family="sans", color="black"), #stylisation
        axis.text = element_text(size=9, color="black"),
        axis.title=element_text(size=9, face = "bold"),
        plot.subtitle=element_text(size=10.5),
        plot.title=element_text(size=11, face = "bold"), 
        legend.position ="none") -> cv.all.1

df_psd.bulk.zfish_atac %>% 
  group_by(celltype) %>% 
  distinct(cv_celltype, cv) %>% 
  ggplot(aes(x=celltype, y=cv_celltype, fill=celltype))+
  geom_col(alpha=0.8, color="gray10")+
  scale_y_continuous(n.breaks=10)+
  theme_bw()+ theme(text = element_text(size = 9, family="sans", color="black"), #stylisation
        axis.text = element_text(size=9, color="black"),
        axis.title=element_text(size=9, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.subtitle=element_text(size=10.5),
        plot.title=element_text(size=11, face = "bold"), 
        legend.position ="none") ->cv.celltype.1

cv.all.1|cv.celltype.1
```


## 3.5 Count distribution by cell type

Then, the count distribution by cell type was plotted analogously to "02-11-ATAC-Seq-Analysis-Step-3.2.1-log 1+x on all celltypes". 
```{r violin plot counts}
df_psd.bulk.zfish_atac %>% 
  mutate(all="all") %>% 
 ggplot( aes( x=all, y=pk_counts))+
  geom_violin(color="grey20", alpha=0.3, fill="gray30", trim=F)+
  coord_fixed(ratio = 20)+
  theme_bw()+ theme(text = element_text(size = 4, family="sans", color="black"), #stylisation
        axis.text = element_text(size=9, color="black"),
        axis.title=element_text(size=9, face = "bold"),
        plot.subtitle=element_text(size=10.5),
        plot.title=element_text(size=11, face = "bold"), 
        legend.position ="none") -> v.plot.all.1


 ggplot(df_psd.bulk.zfish_atac, aes(x=celltype, y=pk_counts, fill = celltype))+
  geom_violin(color="grey20", alpha=0.3, trim=F)+
   coord_equal(ratio = 20)+
  theme_bw()+ theme(text = element_text(size = 4, family="sans", color="black"), #stylisation
        axis.text = element_text(size=9, color="black"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title=element_text(size=9, face = "bold"),
        plot.subtitle=element_text(size=10.5),
        plot.title=element_text(size=11, face = "bold"), 
        legend.position ="none") -> v.plot.cells.1
 
 v.plot.all.1|v.plot.cells.1
```

The data is hardly visible so I made the following adjustments: boxplot and scale limits set to 0 and 0.1

The resulting plot can be seen down below. 
```{r violin plot counts plus boxplot}
df_psd.bulk.zfish_atac %>% 
  mutate(all="all") %>% 
 ggplot( aes( x=all, y=pk_counts))+
  geom_violin(color="grey20", alpha=0.3, fill="gray30", trim=F)+
  geom_boxplot(width = 0.07)+
  stat_boxplot(geom ='errorbar', width=0.2, position=position_dodge(0.5))+
  scale_y_continuous(limits = c(0,0.1))+
  coord_fixed(ratio = 100)+
  theme_bw()+ theme(text = element_text(size = 9, family="sans", color="black"), #stylisation
        axis.text = element_text(size=9, color="black"),
        axis.title=element_text(size=9, face = "bold"),
        plot.subtitle=element_text(size=10.5),
        plot.title=element_text(size=11, face = "bold"), 
        legend.position ="none") -> v.plot.all.box.1


 ggplot(df_psd.bulk.zfish_atac, aes(x=celltype, y=pk_counts, fill = celltype))+
  geom_violin(color="grey20", alpha=0.3, trim=F)+
   geom_boxplot(width = 0.07)+
  stat_boxplot(geom ='errorbar', width=0.2, position=position_dodge(0.5))+
   scale_y_continuous(limits = c(0,0.1))+
  # coord_equal(ratio = 20)+
   #facet_wrap(facets=vars(celltype), scales="free_x")+
  theme_bw()+ theme(text = element_text(size = 9, family="sans", color="black"), #stylisation
        axis.text = element_text(size=9, color="black"),
        #axis.text.x = element_blank(),
        axis.text.x = element_text(angle = 65, hjust = 1),
        axis.title=element_text(size=9, face = "bold"),
        plot.subtitle=element_text(size=10.5),
        plot.title=element_text(size=11, face = "bold"), 
        legend.position ="none") -> v.plot.cells.box.1
 
 v.plot.all.box.1|v.plot.cells.box.1
```

Overall, digestive system cells and erythroid lineage cells are the only cell types, which are not normally distributed. This was also the case in "02-11-ATAC-Seq-Analysis-Step-3.2.1-log 1+x on all celltypes". However, the count values all center around 0.01 be it per cell type or overall. Therefore, the data points were not effectively normalized by seurat's "LogNormalize". For that reason, only manual normalization should be used for further analysis and for assessing the quality of the data set. 

# Session info

Information about the coding environment can be found below.
```{r session info}
sessionInfo()
```



