---
title: "Step 2: Establishing a workflow for ATAC-seq-QC"
subtitle: "Part B (step 2.2): ATAC-seq-data-analysis-QC with manual ln(1+x) normalization"
date: "2025-02-10"
author: Xenia Sirodzha
output:
  html_document: 
    toc: yes
    toc_depth: 4
    toc_float: yes
    df_print: paged
---

Here, I aimed to establish a workflow for ATAC-Seq-QC by performing typical ATAC-seq-QC steps on a reduced set of data. All steps were approved by Alexander Sasse. 

As the normalization options provided for Seurat's `AggregateExpression` function are not sufficently normalizing the ATAC-seq data, in this step, the data was pseudo-bulked using Seurat's `AggregateExpression` function and normalized manually afterwards. To assess the effects of this transformations the data was partially visualized. All plots were analogous to the ones in "02-06-ATAC-Seq-Analysis-Step-1.2.1 seurat norm". Overall, the goal was to determine which normalization method to use ATAC-seq QC. 

# Libraries

These are the packages required for the code below.

```{r libraries}
library(tidyverse) 
library(binom)
library(lubridate)
library(broom)
library(scales) #for scale transformations
library(cowplot) #arranging plots 
library(ggpubr)
library(ggridges)
library(ggthemes) #stylising plots
library(ggplot2)
library(ggsci)
library(wesanderson)
library(readxl) #for loading excel files
library(Seurat) #for rna-seq data, atac-seq data
library(Signac) #for atac seq data
```


# 1. Loading and inspecting 

First, I loaded the sn-ATAC-seq Seurat Object, which contains the count matrix of 5 cell types and 1000 randomly selected peaks. More details about this Seurat Object can be found in "02-06-ATAC-Seq-Analysis-Step-1.1". 
```{r loading and overview of sub_zfish_atac}
#load file

sub_zfish_atac <- readRDS("Data/sub_zfish_atac.rds")
glimpse(sub_zfish_atac)

# overview 

sub_zfish_atac@assays$atac$counts[1:10,1:2]
```

Overall, this Seurat object contains both the count matrix and the meta data. 

# 2. Data transformation
As pseudobulking using Seurat's `AggregateExpression` function is functional (more details in "02-06-ATAC-Seq-Analysis-Step-1.2.1 seurat norm"), I used this function for this operation. Then, I normalized manually with a ln(1+x) transformation by using my own code. 
To ensure that all non determinant code is reproducible, I set a seed.

```{r setting seed for reproducibility}
## ensuring always the same random result
set.seed(59762489)
 
#testing if setting the seed worked
sample(1000,1) #running multiple times, same result
```

## 2.1 Pseudo bulking and normalisation

To generate a matrix, which contains all un-normalized counts per cell type, I pseudo-bulked the count data from the `sub_zfish_atac` object by setting the arguments to: `return.seurat=F` and no specified normalization method, which is the default of this function.

```{r pseudo-bulking}
psd.bulk.sub_atc.lg1x= AggregateExpression(sub_zfish_atac, group.by = "celltype")
```

During ln(1+x) normalization, to calculate the average counts per 1000 cells before applying the ln transformation (`log` function in R) , all counts need to be divided by the number of cells in each celltype. To achieve this I performed the steps below.

First, I extracted the number of cells for each celltype. 
```{r extraction of number of cells per celltype}
sub_zfish_atac@meta.data %>% 
  group_by(celltype) %>% 
  tally() %>% #extracts
  pivot_wider(names_from = celltype, values_from = n) %>% 
  as.matrix() -> n_cells
n_cells
```

Then, I tested whether dividing the counts by the number of cells within each cell type would work with the loop down below on a small scale.

```{r testing the loop on a small scale}
psd.bulk.sub_atc.lg1x$atac[1:5,1:5] -> test
test

## Loop through columns in n_cells
for (i in seq_along(colnames(n_cells))) {
  ct <- colnames(n_cells)[i]  # Extract column name from n_cells
  
  ## Loop through columns in test
  for (j in seq_along(colnames(psd.bulk.sub_atc.lg1x$atac[1:5,1:5]))) {
    atc <- colnames(test)[j]  # Extract column name
    
    ## If names match divide and store in test
    if (atc == ct) {
      test[,j] <- (test[, j] / n_cells[, i])*1000
      break  # stop if everything is matched 
    }
  }
}

test

```

The loop was suitable for the operation. So i proceeded with the operations below.

```{r performing the loop}
#Dividing the counts by the number of cells in each cell type

psd.bulk.sub_atc.lg1x$atac[1:5,1:5] #input

for (i in seq_along(colnames(n_cells))) {
  ct <- colnames(n_cells)[i]  
  

  for (j in seq_along(colnames(psd.bulk.sub_atc.lg1x$atac))) {
    atc <- colnames(psd.bulk.sub_atc.lg1x$atac)[j]  
    

    if (atc == ct) {
      psd.bulk.sub_atc.lg1x$atac[,j] <- (psd.bulk.sub_atc.lg1x$atac[, j] / n_cells[, i])*1000
      break  # stop if everything is matched 
    }
  }
}

psd.bulk.sub_atc.lg1x$atac[1:5,1:5] #output

```

Then to log transform, I tested the operation on a small scale and proceeded afterwards.

```{r testing log transformation on a small scale}
#Test what the operation does on a small scale

psd.bulk.sub_atc.lg1x$atac[1:3,1:5] #input

log(psd.bulk.sub_atc.lg1x$atac[1:3,1:5]+1) #output
```

I transformed the data in the manner shown below.

```{r performing log transformation}
psd.bulk.sub_atc.lg1x=log(psd.bulk.sub_atc.lg1x$atac+1)

psd.bulk.sub_atc.lg1x[1:10,1:5]
```


## 2.2 Data wrangling for plotting by celltype

After ln(1+x) normalization, I wrangled the data to be tidyverse ggplot compatible. First, I converted the matrix to a data frame, then I calculated the values which will be plotted as well(sum of pk_counts, coefficient of variance (cv)).

```{r data wrangling}
as.data.frame(psd.bulk.sub_atc.lg1x)%>% 
  pivot_longer(cols=!contains("rownames"), ,values_to = "pk_counts", names_to = "celltype") %>% 
  mutate(pk_count_sum= sum(pk_counts)) %>% 
  mutate(cv=sd(pk_counts) / mean(pk_counts)) %>% 
  group_by(celltype) %>% 
  mutate(pk_count_sum_celltype= sum(pk_counts)) %>% 
  mutate(cv_celltype=sd(pk_counts) / mean(pk_counts)) %>% 
  mutate(all="all celltypes")-> df_psd.bulk.sub_atc.lg1x

df_psd.bulk.sub_atc.lg1x
```


# 3 Visualization

Overall, all data was plotted analogically to the visualization in "02-06-ATAC-Seq-Analysis-Step-1.2.1 seurat norm". 

## 3.1 Distribution after pseudo-bulking by celltype

Counts were plotted in a histogram overall and then by cell type. 
```{r histograms counts}
# Overall

df_psd.bulk.sub_atc.lg1x %>% 
 ggplot(aes(x=pk_counts))+geom_histogram(binwidth = 0.2, color="gray15", alpha=0.8)+
 scale_y_continuous(breaks = scales::breaks_pretty(n=10))+
  coord_fixed(ratio = 0.03)+
  theme_bw()+ theme(text = element_text(size = 9, family="sans", color="black"), #stylisation
        axis.text = element_text(size=9, color="black"),
        axis.title=element_text(size=9, face = "bold"),
        plot.subtitle=element_text(size=10.5),
        plot.title=element_text(size=11, face = "bold"), 
        legend.position ="none") ->dis.counts.1


# by cell type 

 ggplot(df_psd.bulk.sub_atc.lg1x, aes(x=pk_counts, fill = celltype))+
  geom_histogram(binwidth = 0.2, color="gray15", alpha=0.8)+
  scale_y_continuous(breaks = scales::breaks_pretty(n=10))+
  facet_wrap(facets = "celltype")+
  theme_bw()+ theme(text = element_text(size = 9, family="sans", color="black"), #stylisation
        axis.text = element_text(size=9, color="black"),
        axis.title=element_text(size=9, face = "bold"),
        plot.subtitle=element_text(size=10.5),
        plot.title=element_text(size=11, face = "bold"), 
        legend.position ="none") ->dis.counts.cell.1

# display

dis.counts.1|dis.counts.cell.1
```

Generally, counts are somewhat normally distributed. There are no cell type specific trends. Most counts are between the values 1 and 5 which can be typical for well normalized ATAC-seq-data.

## 3.2 Sum of all peaks

I counted and visualized the sum of all peaks by cell type and in general.

In addition, I extracted the value for the sum of all peaks. It is  12770.74. Which is higher than with either Seurat provided normalization method (see 02-06-ATAC-Seq-Analysis-Step-1.2.1 seurat norm). 
```{r extracting value sum of all peaks}
unique(df_psd.bulk.sub_atc.lg1x$pk_count_sum)
```

The sums of all peaks were visualized in a barplot.
```{r bar plot sum of all peaks}

# Overall

df_psd.bulk.sub_atc.lg1x %>% 
  ungroup() %>% 
  distinct(all, pk_count_sum) %>% 
  ggplot(aes(x=all, y=pk_count_sum))+
  geom_col(alpha=0.8, color="gray10")+
  coord_fixed(ratio = 0.0002)+
  theme_bw()+ theme(text = element_text(size = 9, family="sans", color="black"), #stylisation
        axis.text = element_text(size=9, color="black"),
        axis.title=element_text(size=9, face = "bold"),
        plot.subtitle=element_text(size=10.5),
        plot.title=element_text(size=11, face = "bold"), 
        legend.position ="none") -> sum.peaks.all.1


# by cell type

df_psd.bulk.sub_atc.lg1x %>% 
  group_by(celltype) %>% 
  distinct(pk_count_sum_celltype, pk_count_sum) %>% 
  ggplot(aes(x=celltype, y=pk_count_sum_celltype, fill=celltype))+
  geom_col(alpha=0.8, color="gray10")+
  theme_bw()+ theme(text = element_text(size = 9, family="sans", color="black"), #stylisation
        axis.text = element_text(size=9, color="black"),
        axis.title=element_text(size=9, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.subtitle=element_text(size=10.5),
        plot.title=element_text(size=11, face = "bold"), 
        legend.position ="none") -> sum.peaks.celltype.1

sum.peaks.all.1|sum.peaks.celltype.1
```

Apart from the YSL/presumptive endoderm cells with a little over 3000 peak count sum, all cells have a similar peak count sum. Overall, the proportion of the sum of all peaks is similar to previous normalization methods. 

## 3.3 Coefficient of variation 

The coefficient of variation was visualized as the sum of all peaks previously.
```{r bar plot coefficient of variation}

# Overall

df_psd.bulk.sub_atc.lg1x %>% 
  ungroup() %>% 
  distinct(all, cv) %>% 
  ggplot(aes(x=all, y=cv ))+
  geom_col(alpha=0.8, color="gray10")+
  scale_y_continuous(n.breaks=10)+
  coord_fixed(ratio = 6)+
  theme_bw()+ theme(text = element_text(size = 9, family="sans", color="black"), #stylisation
        axis.text = element_text(size=9, color="black"),
        axis.title=element_text(size=9, face = "bold"),
        plot.subtitle=element_text(size=10.5),
        plot.title=element_text(size=11, face = "bold"), 
        legend.position ="none") -> cv.all.1


# by cell type

df_psd.bulk.sub_atc.lg1x %>% 
  group_by(celltype) %>% 
  distinct(cv_celltype, cv) %>% 
  ggplot(aes(x=celltype, y=cv_celltype, fill=celltype))+
  geom_col(alpha=0.8, color="gray10")+
  scale_y_continuous(n.breaks=10)+
  theme_bw()+ theme(text = element_text(size = 9, family="sans", color="black"), #stylisation
        axis.text = element_text(size=9, color="black"),
        axis.title=element_text(size=9, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.subtitle=element_text(size=10.5),
        plot.title=element_text(size=11, face = "bold"), 
        legend.position ="none") ->cv.celltype.1

cv.all.1|cv.celltype.1
```

Overall, coefficients of variation of each cell type are very similar to one another and to the coefficient of variation of all cells. However, YSL/presumptive endoderm cells have the lowest coefficient of variation. 

## 3.4 Count distribution by cell type

Then, just as previously count distribution was visualized with a violin plot.

```{r violin plot counts}
df_psd.bulk.sub_atc.lg1x %>% 
  mutate(all="all") %>% 
 ggplot( aes( x=all, y=pk_counts))+
  geom_violin(color="grey20", alpha=0.3, fill="gray30", trim=F)+
  coord_fixed(ratio = 0.3)+
  theme_bw()+ theme(text = element_text(size = 9, family="sans", color="black"), #stylisation
        axis.text = element_text(size=9, color="black"),
        axis.title=element_text(size=9, face = "bold"),
        plot.subtitle=element_text(size=10.5),
        plot.title=element_text(size=11, face = "bold"), 
        legend.position ="none") -> v.plot.all.1


 ggplot(df_psd.bulk.sub_atc.lg1x, aes(x=celltype, y=pk_counts, fill = celltype))+
  geom_violin(color="grey20", alpha=0.3, trim=F)+
  theme_bw()+ theme(text = element_text(size = 9, family="sans", color="black"), #stylisation
        axis.text = element_text(size=9, color="black"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title=element_text(size=9, face = "bold"),
        plot.subtitle=element_text(size=10.5),
        plot.title=element_text(size=11, face = "bold"), 
        legend.position ="none") -> v.plot.cells.1
 
 v.plot.all.1|v.plot.cells.1
```

To visualize median and quantiles, a boxplot was added. 
```{r violin plot counts plus boxplot}
df_psd.bulk.sub_atc.lg1x %>% 
  mutate(all="all") %>% 
 ggplot( aes( x=all, y=pk_counts))+
  geom_violin(color="grey20", alpha=0.3, fill="gray30", trim=F)+
  geom_boxplot(width = 0.07)+
  stat_boxplot(geom ='errorbar', width=0.2, position=position_dodge(0.5))+
  scale_y_continuous(breaks = seq(0,10,1))+ #y-axis more easily readable
  coord_fixed(ratio = 0.3)+
  theme_bw()+ theme(text = element_text(size = 9, family="sans", color="black"), #stylisation
        axis.text = element_text(size=9, color="black"),
        axis.title=element_text(size=9, face = "bold"),
        plot.subtitle=element_text(size=10.5),
        plot.title=element_text(size=11, face = "bold"), 
        legend.position ="none") -> v.plot.all.box.1


 ggplot(df_psd.bulk.sub_atc.lg1x, aes(x=celltype, y=pk_counts, fill = celltype))+
  geom_violin(color="grey20", alpha=0.3, trim=F)+
  geom_boxplot(width = 0.07)+
  stat_boxplot(geom ='errorbar', width=0.2, position=position_dodge(0.5))+
    scale_y_continuous(breaks = seq(0,10,1))+ #y-axis more easily readable
  theme_bw()+ theme(text = element_text(size = 9, family="sans", color="black"), #stylisation
        axis.text = element_text(size=9, color="black"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title=element_text(size=9, face = "bold"),
        plot.subtitle=element_text(size=10.5),
        plot.title=element_text(size=11, face = "bold"), 
        legend.position ="none") -> v.plot.cells.box.1
 
 v.plot.all.box.1|v.plot.cells.box.1
```

The counts are similarly distributed, however values are more centered around count value 2.5. Most counts have values ranging from 0.5 to 4. In addition, YSL/presumptive endoderm cells have higher counts than all other cell types. In contrast to the normalized data  in "02-06-ATAC-Seq-Analysis-Step-1.2.1 seurat norm", most counts range from 0.5 to 5, which is similar to what one would expect in well-normalized ATAC-seq data. 

# 4. Summary

Overall, manual ln(1+x) normalization after pseudo-bulking, leads to normally distributed counts with a wider range than in with Seurat's `AggregateExpression` function normalized data. In addition, the values are centered around a higher value 2.5. Therefore, it is likely that this the normalization method I should use when I perform ATAC-seq QC. Nonetheless, both log based normalization methods were applied to the entire data set after pseudobulking. 


# Session info

Information about the coding environment can be found below.
```{r session info}
sessionInfo()
```
