---
title: "Step 4: Preparation of the sn-ATAC-seq-data set for modelling by Alexander Sasse's group"
subtitle: "Step 4.2: Pseudobulking and normalizing subsetted celltypes"
date: "2025-03-06"
author: Xenia Sirodzha
output:
  html_document: 
    toc: yes
    toc_depth: 4
    toc_float: yes
    df_print: paged
---

# Libraries 

These are the packages required for the coding in this document.
```{r libraries}
library(tidyverse) 
library(binom)
library(broom)
library(Seurat) #for rna-seq data, atac-seq data
library(Signac) #for atac seq data
```

# Data loading

First, the data needed for normalization was loaded. Its origin is further explained in "03-06-ATAC-Seq-Analysis-Step-4.1-subset-celltypes".
```{r data loading}
readRDS("Data/psd.bulk.zfish_atac.10hpf.rds") -> psd.bulk.zfish_atac.10hpf
readRDS("Data/zfish_atac.10hpf.rds") ->zfish_atac.10hpf
```

# Data transformation

Overall, the data was transformed analogous to "02-11-ATAC-Seq-Analysis-Step-3.1.1-log 1+x on all celltypes".

First, the number of cells was extracted.
```{r extracting number of cells}
zfish_atac.10hpf@meta.data %>% 
  group_by(celltype) %>% 
  tally() %>% 
  pivot_wider(names_from = celltype, values_from = n) %>% 
  as.matrix() -> n_cells
n_cells
```


Then, I employed the following loop to extract the correct number of cells of each cell type to divide the counts of the matrix containing the pseudo-bulk with. 
```{r manual normalization}
#Dividing the counts by the number of cells in each cell type

psd.bulk.zfish_atac.10hpf$atac[1:5,1:5] #input

for (i in seq_along(colnames(n_cells))) {
  ct <- colnames(n_cells)[i]  #extracting cell type names
  
#loop along cell type names in the pseudo bulk matrix
  for (j in seq_along(colnames(psd.bulk.zfish_atac.10hpf$atac))) {
    atc <- colnames(psd.bulk.zfish_atac.10hpf$atac)[j]  
    
#if the colnames match divide by the number of cells
    if (atc == ct) {
      psd.bulk.zfish_atac.10hpf$atac[,j] <- (psd.bulk.zfish_atac.10hpf$atac[, j] / n_cells[, i])*1000
      break  # stop if everything is matched 
    }
  }
}

psd.bulk.zfish_atac.10hpf$atac[1:5,1:5] #output

```

After that, the matrix was ln(1+x) transformed in the manner seen below.

```{r manual ln transformation}
psd.bulk.zfish_atac.10hpf.lg1x=log(psd.bulk.zfish_atac.10hpf$atac+1)

psd.bulk.zfish_atac.10hpf.lg1x[1:10,1:5] #output
```

Lastly, it was saved as a csv file for easier readability and further use.
```{r saving as csv, eval=FALSE}
#save as file

write.csv(psd.bulk.zfish_atac.10hpf.lg1x, file = "Data/psd.bulk.zfish_atac.10hpf.lg1x.csv")
```

# Session info 

```{r session info}
sessionInfo()
```
